Optimizing Abstract Abstract Machines

### 2012-11-15 paper/paper.tex
Link to implementation.
ispell.

\begin{figure}[t]
\begin{center}
\begin{tabular}{ccc}
\raisebox{1ex-\height}{
\includegraphics[height=3.5in]{introspective-base.pdf}}
&
\raisebox{1ex-\height}{
\includegraphics[height=3.5in]{introspective-lazy.pdf}}
&
\raisebox{1ex-\height}{
\includegraphics[height=3in]{introspective-lazyc.pdf}}
\\
(a) Baseline
&
(b) Lazy
&
(c) Compiled (\& lazy)
\end{tabular}
\end{center}
\caption{Example state graphs for the program above.  Part (a) shows
  the result of the baseline analyzer.  It has long ``corridor''
  transitions and ``diamond'' subgraphs that fan-out from
  nondeterminism and fan-in from joins.  Part (b) shows the result of
  performing nondeterminism lazily and thus avoids many of the diamond
  subgraphs.  Part (c) shows the result of abstract compilation that
  removes intepretive overhead in the form of intermediate states,
  thus minimizing the corridor transitions.  The end result is a more
  compact abstraction of the program that can be generated faster.}
\label{fig:state-graphs}
\end{figure}

@@

Section~\ref{sec:eval} describes the evaluation of each optimiztion
technique applied to an implementation supporting a more realistic set
of features, including mutation, first-class control, compound data, a
full numeric tower and many more forms of primitive data and
operations.
%
We evaluate this implementation against a set of benchmark programs
drawn from the literature.
%
For all benchmarks, the optimized analyzer outperforms the baseline
by at least a factor of
% 475
two to
% 4,382
three orders of magnitude.

@@

The semantics are defined in terms of a machine model.  The machine
compoents are defined in figure~\ref{fig:domains} and
figure~\ref{fig:aam} defines the transition relation.  The evaluation
of a program is defined as the set of states reachable by the
reflexive, transitive closure of the machine transition relation.  The
machine is a very slight variation on a standard abstract machine for
ISWIM in ``eval, continue, apply'' form~\cite{dvanhorn:Danvy:DSc}.  It
can be systematically derived from a definitional interpreter through
a continuation-passing style transformation and defunctionalization,
or from a structural operational semantics using the refocusing
construction of Danvy and
Nielsen~\cite{dvanhorn:Danvy-Nielsen:RS-04-26}.

@@

\paragraph{Concrete interpretation} To characterize concrete interpretration, set the implicit
parameters of the relation given in figure~\ref{fig:aam} as follows:
\begin{align*}
\mathit{push}_\mlab^\mcntr(\msto,\mkont) &= \maddr,\msto\sqcup[\maddr\mapsto\{\mkont\}]
\mbox{ where }\maddr \notin\msto
\\
\mathit{bind}_\mlab^\mcntr(\menv,\msto,\mvar,\mval) &= \menv[\mvar\mapsto\maddr],\msto\sqcup[\maddr\mapsto\{\mval\}]
\mbox{ where }\maddr \notin\msto
\end{align*}
The resulting relation is non-deterministic in its choice of
addresses, however it must always choose a fresh address when
allocating a continuation or variable binding.  If we consider machine
states equivalent up to consistent renaming, this relation defines
a deterministic machine.  (The relation is really a function.)

@@

All that remains is the interpretation of primitives.  For abstract
intepretation, we set $\interpdelta$ to the function that returns
$\sNum$ on all inputs---a symbolic value we interpret as denoting the
set of all integers.

@@

We can cast this exploration process in terms of a fixed-point calculation.
%
Given the initial state $\varsigma_0$ and the transition relation $\machstep$,
we define the global transfer function:
\begin{equation*}
 F_{\varsigma_0} : \mathcal{P}(\mathit{State}) \to \mathcal{P}(\mathit{State})
 \text.
\end{equation*}
Internally, this global transfer function computes the successors of all supplied states, and then includes the inital state:
\begin{equation*}
  F_{\varsigma_0}(S) = \{ \varsigma_0 \} \cup \{ \varsigma' \mathrel{|} \varsigma \in S \text{ and } \varsigma \machstep \varsigma' \}\text. 
\end{equation*}
Then, the evaluator for the analysis computes the least fixed-point of the global transfer function:
\begin{equation*}
 \eval(e) = \mathrm{lfp}(F_{\varsigma_0})\text{,}
\end{equation*}
where $\varsigma_0 = \ev[^\mtcntr]{e, \varnothing, \varnothing, \kmt}$.

@@

In the next subsection, we'll fix this with a widening and reach polynomial
(albeit of a high degree) complexity.
%
This widening effictively lifts the store out of individual states to create
a single, global shared store for all.

@@

A common technique to accelerate convergence in flow analyses is to share a
common, global store.
%
To retain soundness, this store grows monotonically.
%
Formally, we can cast this optimization as a second abstraction or as the
application of a widening operator during the fixed-point iteration.
%
In the ISWIM language, such a widening makes 0-CFA quartic in the size of the
program.
%
Thus, in one step, complexity drops from intractable exponentiality to merely
daunting polynomiality.

@@

\paragraph{Option 2: Regain precision; complicate implementation}
The administrative nature of laziness means that we could remove the
loss in precision by duplicating the reduction relation to specialize
variable lookup. This works since in the semantics of ISWIM with
store-allocated results consumes the value component of states in one
step. This is not the case for semantics that replicate the value
component across reductions, say for popping off exception handler
frames. Further convolution is needed to remove the administrative
nature of laziness in these semantics. Due to the increase of
conceptual complexity for negligable benefit, we decided against this
approach.

@@

The abstraction of compound data structures like lists has deep implications 
for precision and performance.
%
The classic list-heavy benchmark {\tt boyer}---a simplified implementation
of the Boyer-Moore theorem-prover---drove our own considerations for
abstractions for compound data structures.
%
Boyer's fluent use of literal list dooms the ``natural'' abstraction
of lists to fail from overprecision.
%
In short, if we interpret a literal list {\tt '(a b c)} as {\tt (cons 'a (cons 'b
(cons 'c '())))}, even a monovariant allocation stategy for abstract
cons cells will precisely and exactly create a 3-cell list.

@@

Upon evaluating a {\tt cons} application, we instead allocate two
addresses $a$ and $d$, join them to the respective values in the
store, and return the flattened {\tt (cons $a$ $d$)} value. Since
these addresses are all distinguished at different syntactic callsites
in the uniform $k$-CFA allocation strategy, and quoted lists are sugar
for a sequence of calls to {\tt cons}, this abstraction explodes the
value space. Analyzing a function that counts the number of atoms in a
literal s-expression would actually interpret that function at least
that number of times (more because of intermediate conses). Indeed,
even in our fastest implementation, this abstraction causes the
analysis of Boyer to be 430 times slower than the approach we will now
describe.

@@

@@

The only analyzers we were able to find that proved capable of
analyzing the full suite of benchmarks considered here were the Soft
Typing system of Wright and
Cartwright~\cite{dvanhorn:Wright1997Practical} and, in many ways its
successor, the Polymorphic splitting system of Wright and
Jagannathan~\cite{dvanhorn:wright-jagannathan-toplas98}.\footnote{This
  is not a coincidence; these papers set a high standard for
  evaluation, which we consciously aimed to approach.}  Unfortunately, these
analyses compute an inherently different and uncomparable form of
analysis.  Consequently, we have omitted a complete comparison with
these implementations,.  The AAM approach provides more precision in
terms of temporal-ordering of program states, which comes at a cost
that can be avoided in constraint-based approaches.  Consequently
implementation techniques cannot be ``ported'' between these two
approaches.  However, our optimized implementation is within an order
of magnitude of the performance of Wright and Jaganathan's analyzer.
Although we would like to improve this to be more competitive, the
optimized AAM approach still has many strengths to recommend it in
terms of precision, ease of implementation and verification, and rapid
design.

>>>>

\begin{figure}[t]
\begin{center}
\begin{tabular}{ccc}
\raisebox{1ex-\height}{
\includegraphics[height=3.5in]{introspective-base.pdf}}
&
\raisebox{1ex-\height}{
\includegraphics[height=3.5in]{introspective-lazy.pdf}}
&
\raisebox{1ex-\height}{
\includegraphics[height=3in]{introspective-lazyc.pdf}}
\\
(a) Baseline
&
(b) Lazy
&
(c) Compiled (\& lazy)
\end{tabular}
\end{center}
\caption{Example state graphs for the program above.  Part (a) shows
  the result of the baseline analyzer.  It has long ``corridor''
  transitions and ``diamond'' subgraphs that fan-out from
  nondeterminism and fan-in from joins.  Part (b) shows the result of
  performing nondeterminism lazily and thus avoids many of the diamond
  subgraphs.  Part (c) shows the result of abstract compilation that
  removes interpretive overhead in the form of intermediate states,
  thus minimizing the corridor transitions.  The end result is a more
  compact abstraction of the program that can be generated faster.}
\label{fig:state-graphs}
\end{figure}

@@

Section~\ref{sec:eval} describes the evaluation of each optimization
technique applied to an implementation supporting a more realistic set
of features, including mutation, first-class control, compound data, a
full numeric tower and many more forms of primitive data and
operations.
%
We evaluate this implementation against a set of benchmark programs
drawn from the literature.
%
For all benchmarks, the optimized analyzer outperforms the baseline
by at least a factor of
% 475
two to
% 4,382
three orders of magnitude.

@@

The semantics are defined in terms of a machine model.  The machine
components are defined in figure~\ref{fig:domains} and
figure~\ref{fig:aam} defines the transition relation.  The evaluation
of a program is defined as the set of states reachable by the
reflexive, transitive closure of the machine transition relation.  The
machine is a very slight variation on a standard abstract machine for
ISWIM in ``eval, continue, apply'' form~\cite{dvanhorn:Danvy:DSc}.  It
can be systematically derived from a definitional interpreter through
a continuation-passing style transformation and defunctionalization,
or from a structural operational semantics using the refocusing
construction of Danvy and
Nielsen~\cite{dvanhorn:Danvy-Nielsen:RS-04-26}.

@@

\paragraph{Concrete interpretation} To characterize concrete interpretation, set the implicit
parameters of the relation given in figure~\ref{fig:aam} as follows:
\begin{align*}
\mathit{push}_\mlab^\mcntr(\msto,\mkont) &= \maddr,\msto\sqcup[\maddr\mapsto\{\mkont\}]
\mbox{ where }\maddr \notin\msto
\\
\mathit{bind}_\mlab^\mcntr(\menv,\msto,\mvar,\mval) &= \menv[\mvar\mapsto\maddr],\msto\sqcup[\maddr\mapsto\{\mval\}]
\mbox{ where }\maddr \notin\msto
\end{align*}
The resulting relation is non-deterministic in its choice of
addresses, however it must always choose a fresh address when
allocating a continuation or variable binding.  If we consider machine
states equivalent up to consistent renaming, this relation defines
a deterministic machine.  (The relation is really a function.)

@@

All that remains is the interpretation of primitives.  For abstract
interpretation, we set $\interpdelta$ to the function that returns
$\sNum$ on all inputs---a symbolic value we interpret as denoting the
set of all integers.

@@

We can cast this exploration process in terms of a fixed-point calculation.
%
Given the initial state $\varsigma_0$ and the transition relation $\machstep$,
we define the global transfer function:
\begin{equation*}
 F_{\varsigma_0} : \mathcal{P}(\mathit{State}) \to \mathcal{P}(\mathit{State})
 \text.
\end{equation*}
Internally, this global transfer function computes the successors of all supplied states, and then includes the initial state:
\begin{equation*}
  F_{\varsigma_0}(S) = \{ \varsigma_0 \} \cup \{ \varsigma' \mathrel{|} \varsigma \in S \text{ and } \varsigma \machstep \varsigma' \}\text. 
\end{equation*}
Then, the evaluator for the analysis computes the least fixed-point of the global transfer function:
\begin{equation*}
 \eval(e) = \mathrm{lfp}(F_{\varsigma_0})\text{,}
\end{equation*}
where $\varsigma_0 = \ev[^\mtcntr]{e, \varnothing, \varnothing, \kmt}$.

@@

In the next subsection, we'll fix this with a widening and reach polynomial
(albeit of a high degree) complexity.
%
This widening effectively lifts the store out of individual states to create
a single, global shared store for all.

@@

A common technique to accelerate convergence in flow analyses is to share a
common, global store.
%
To retain soundness, this store grows monotonically.
%
Formally, we can cast this optimization as a second abstraction or as the
application of a widening operator during the fixed-point iteration.
%
In the ISWIM language, such a widening makes 0-CFA quartic in the size of the
program.
%
Thus, in one step, complexity drops from intractable exponentially to a merely
daunting polynomial.

@@

\paragraph{Option 2: Regain precision; complicate implementation}
The administrative nature of laziness means that we could remove the
loss in precision by duplicating the reduction relation to specialize
variable lookup. This works since in the semantics of ISWIM with
store-allocated results consumes the value component of states in one
step. This is not the case for semantics that replicate the value
component across reductions, say for popping off exception handler
frames. Further convolution is needed to remove the administrative
nature of laziness in these semantics. Due to the increase of
conceptual complexity for negligible benefit, we decided against this
approach.

@@

The abstraction of compound data structures like lists has deep implications 
for precision and performance.
%
The classic list-heavy benchmark {\tt boyer}---a simplified implementation
of the Boyer-Moore theorem-prover---drove our own considerations for
abstractions for compound data structures.
%
Boyer's fluent use of literal list dooms the ``natural'' abstraction
of lists to fail from over precision.
%
In short, if we interpret a literal list {\tt '(a b c)} as {\tt (cons 'a (cons 'b
(cons 'c '())))}, even a monovariant allocation strategy for abstract
cons cells will precisely and exactly create a 3-cell list.

@@

Upon evaluating a {\tt cons} application, we instead allocate two
addresses $a$ and $d$, join them to the respective values in the
store, and return the flattened {\tt (cons $a$ $d$)} value. Since
these addresses are all distinguished at different syntactic call sites
in the uniform $k$-CFA allocation strategy, and quoted lists are sugar
for a sequence of calls to {\tt cons}, this abstraction explodes the
value space. Analyzing a function that counts the number of atoms in a
literal s-expression would actually interpret that function at least
that number of times (more because of intermediate conses). Indeed,
even in our fastest implementation, this abstraction causes the
analysis of Boyer to be 430 times slower than the approach we will now
describe.

@@

Source code of the implementation and benchmark suite is at:

\begin{center}
\url{https://github.com/dvanhorn/oaam}
\end{center}

@@

The only analyzers we were able to find that proved capable of
analyzing the full suite of benchmarks considered here were the Soft
Typing system of Wright and
Cartwright~\cite{dvanhorn:Wright1997Practical} and, in many ways its
successor, the Polymorphic splitting system of Wright and
Jagannathan~\cite{dvanhorn:wright-jagannathan-toplas98}.\footnote{This
  is not a coincidence; these papers set a high standard for
  evaluation, which we consciously aimed to approach.}  Unfortunately, these
analyses compute an inherently different and incomparable form of
analysis.  Consequently, we have omitted a complete comparison with
these implementations,.  The AAM approach provides more precision in
terms of temporal-ordering of program states, which comes at a cost
that can be avoided in constraint-based approaches.  Consequently
implementation techniques cannot be ``ported'' between these two
approaches.  However, our optimized implementation is within an order
of magnitude of the performance of Wright and Jaganathan's analyzer.
Although we would like to improve this to be more competitive, the
optimized AAM approach still has many strengths to recommend it in
terms of precision, ease of implementation and verification, and rapid
design.
